---
title: "Including covariates in occupancy models"
author: "Deon Roos"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    highlight: monochrome
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float: true
    collapsed: false
    smooth_scroll: false
    df_round1_print: paged
---

# The theory

In the previous document we went through a simple example to get a sense of how occupancy models work. We're now going to begin the process of slowly increasing the complexity. To start, we're going to need to include covariates (often called explanatory variables or independent variables or a wide variety of other names - it's a mess).

Let's revisit our simple occupancy model:

$$
z_i \sim Bernoulli(\psi_i)\\
logit(\psi_i) = \beta_0\\
y_{i,j} \sim Bernoulli(p_{i,j} \times z_i)\\
logit(p_{i,j}) = \alpha_0
$$

where the first two formula represent the state model (i.e. are our species present or absent from site $i$), and the last two form the detection model.

What we're going to do in this section is to add in covariates to these equations, starting with the detection model.

The first thing to note is that $y$ is indexed by both $i$ and $j$, where $i$ was the site and $j$ was the survey. That $j$ is important because it allows us to specify covariates that vary not just by site (e.g. site 1 has 20 trees while site 2 has 10 trees) but by survey as well (e.g. on survey 1 in site 1, the temperature was 20 degrees, 15 degrees in survey 2 and 5 degrees in survey 3). This means that anything that was different in one survey to the next can be accounted for and included in the model (so long as we can measure it).

Specifically, these "survey varying covariates" are features that we think may have made us (or whoever or whatever collected the data) to be more or less effective. Using cameras? Well maybe the presence of fog has a big impact on how likely we are to detect elephants. Doing surveys yourself? Well maybe the hour of day that you did the survey had a big impact. The point is, we can include these variables that deal with differences in detection probability.

I'll simulate a new dataset (with a bit more data for us to work with - 64 sites surveyed 3 times each) to show this off. We can say that the detection covariate in this case is rainfall, but note that the values for rainfall will be centred on zero (don't worry about this).

Here's what the data looks like:

```{r}
library(spOccupancy)
set.seed(1234)
dat <- simOcc(J.x = 8, 
              J.y = 8, 
              n.rep = rep(3, times = 8 * 8), 
              beta = c(1), 
              alpha = c(-2, 0.5))
obs <- dat$y
det_cov <- dat$X.p[,,2]
df <- data.frame(
  survey = rep(1:3, each = 64),
  cov = c(det_cov[,1], det_cov[,2], det_cov[,3]),
  y = c(obs[,1], obs[,2], obs[,3])
)

library(ggplot2)
p1 <- ggplot(df) +
  geom_boxplot(aes(x = factor(survey), y = cov)) +
  geom_jitter(aes(x = factor(survey), y = cov),
              width = 0.2, height = 0) +
  labs(x = "Survey", y = "Rainfall") +
  theme_minimal()

p2 <- ggplot(df) +
  geom_jitter(aes(y = factor(y), x = cov),
              alpha = 0.4, width = 0, height = 0.1) +
  labs(x = "Rainfall", y = "Elephant detection") +
  theme_minimal()

library(patchwork)
p1 + p2 + plot_annotation(tag_levels = "A", tag_suffix = ")")
```

Notice anything? In **A)** we can see that in each survey, there's a lot of variation across the 64 sites in how much rainfall there was; maybe peaking in survey 2. In **B)**, it looks like it might be more likely that we detect elephants more often when there's more rainfall.

Are you sure there's an influence? If so, how strong is it? *Exactly* how strong is it?

There's no way you can answer those questions. That's where we need stats. So let's add in rainfall to our model. I'll recycle the code from the previous document and include rainfall. Here's how we do that.

# Data preparation

Just like in the previous document, we need to include our different datasets into a `list`. If you're still not sure what a `list` is in `R`, think of it like a folder on your computer. You can add lots of different files to a folder, but they're all "tied" together by being within the same folder. That's the same as a `list` in `R`.

Here, I'm going to do something seemingly strange. I'm going to create a `list` and include this into our `etosha` `list`. The reason is that we might have more than one detection covariate, so having a list, though redundant here, will make adding additional variables easier in the future.

```{r}
# Note you wouldn't need to do the dat$X.p[,,2] bit
# That's just because the data is simulated.
det.covs <- list(rain = dat$X.p[,,2])

etosha <- list(
  y = dat$y,
  det.covs = det.covs
)
```

And we're good to go on to the modelling.

# Fitting the model

```{r}
fit <- PGOcc(
  # The state model (i.e. what % that elephants are present?)
  # ~ 1 means we want an intercept only model (no covariates)
  occ.formula = ~ 1, 
  # The observation model (i.e. what % that we see elephants if present?)
  # ~ 1 means the same as above - intercept only
  det.formula = ~rain, 
  # Our carefully formatted dataset
  data = etosha, 
  
  # Details to get the machinery to run that we'll ignore for now
  n.chains = 4,
  n.samples = 2000,
  n.burn = 200,
  verbose = FALSE)
```

```{r}
summary(fit)
```

